MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments

### 2022-03-17 main/MSstatsPTM_Main.tex
main changes

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of Bio. &  Data \\
&  & Design & Conditions & Replicates & availability \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset in this manuscript.
``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository.  \todo{All the experiments were conducted in data-dependent acquisition (DDA) mode.}
}
\label{tab:dataDescription}
\end{table}

@@

The simulation provides an idealistic case. \todo{generated (how many?)  repeated synthetic datasets?}. \todo{I have a hard time following this section. Would it be possible to rewrite the description, by explaining (1) how many repeated datasets, (2) how many proteins per dataset; (3) how many PTMs per protein, (4) how many modified/unmodified features per protein? Then talk about the quantitative aspects: systematic fold change; random variation; where the parameters of the simulation were inspired from. It will help relate this simulation to real data} had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein. Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein. All differential PTMs were generated with an expected log fold change of .75 between conditions. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.
Further details can be found in Supplementary Sec. \ref{supp-sec:dataset1}.

@@

The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments. In this simulation, the feature counts and the proportion of missing values were as observed \todo{on average over all the experimental datasets in this manuscript?}. PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. \todo{There will likely be a comment that missing values are more common among low-abundance features. Did the simulation account for that?} Further details can be found in Supplementary Sec. \ref{supp-sec:dataset2}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The proposed approach corrects \todo{starting here, put all the section titles in past tense} for high FDR in computer simulations}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The proposed approach outperforms existing methods in simulations with missing values and low features}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In a label-free benchmark experiment the proposed approach corrects for bias and outperforms existing methods}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{MSstatsPTM corrects for bias in group comparison TMT experiment}

@@

\subsubsection*{Proposed approach removes confounding in time series TMT experiment}

@@

\subsubsection*{Noisy PTM measurements benefit from additional biological replicates}

>>>>

%%%% Summary of controlled mixtures
\begin{table}[h!]
\centering
\begin{tiny}
\begin{tabular}{|c|c|ccc|c|}
\hline
%%%% colnames of table
& \multirow{2}{*}{Dataset} & Experimental & No. of & No. of Bio. &  Data \\
&  & Design & Conditions & Replicates & availability \\
\hline
\hline
 &&&&& \\[-0.05in]
%%%% DDA:ControlledMix
\multirow{5}{*}{\rotatebox[origin=c]{90}{Known}}  \multirow{5}{*}{\rotatebox[origin=c]{90}{Ground}} \multirow{5}{*}{\rotatebox[origin=c]{90}{Truth}} & Computer simulation 1 - Label-free & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github}\\
 &&&&& \\%[-0.05in]
%\hline 
%&&&&&&&& \\[-0.05in]
% \cline{2-9}
% &&&&&&&& \\[-0.05in]
%%% DDA iPRG2015
& Computer simulation 2 - missing and low features & Group comparison & 2/3/4 & 2/3/5/10 & \href{https://github.com/devonjkohler/MSstatsPTM_simulations}{Github} \\
 &&&&& \\%[0.05in]
%\hline 
% &&&&&& \\[-0.05in]
%%% DDA Cox2014
& SpikeIn benchmark - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/private-dataset.jsp?task=c4c583ecf7f941cdac87f7a4f872517b}{MSV000088971}\\
 &&&&& \\%[-0.05in]
\hline
\multicolumn{6}{c}{ } \\ [0.02in]
\hline 
 &&&&& \\%[-0.05in]
%%%%%%%%%%%%%%%%%%
%%%% biological study
%%%%%%%%%%%%%%%%%%
%% DDA Meierhofer
\multirow{3}{*}{\rotatebox[origin=c]{90}{Biological}} \multirow{3}{*}{\rotatebox[origin=c]{90}{Investigation}} & Human - Ubiquitination - 1mix-TMT & Group comparison & 6 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=b6f0c74c234247678fb0888c6df1f225}{MSV000088966} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Picotti 2009
& Mouse - Phosphorylation - 2mix-TMT & Time course & 6 & 4 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=4878d777c6b34cf8aaf8477e93140c4d}{MSV000085565} \\
&&&&& \\%[-0.05in]
% \cline{2-9}
% &&&&&&& \\[-0.05in]
%%%% SRM Surinova 2015, training
& Human - Ubiquitination - Label-free & Group comparison & 4 & 2 & \href{https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=1b516164de5345108b40b75147dd58b5}{MSV000078977}\\ [0.02in]
 &&&&& \\
 \hline
\end{tabular}\\
\end{tiny}
\caption{ \small {\bf Simulated and experimental datasets in this manuscript} 
```Dataset'' is the code name of the dataset in this manuscript.
``Data availability'' shows the ID of the MassIVE.quant repository or the GitHub repository. All the experiments were conducted in data-dependent acquisition (DDA) mode.
}
\label{tab:dataDescription}
\end{table}

@@

The simulation provides an idealistic case. 24 synthetic datasets were generated exhibiting scenarios with different experimental designs and biological variation. 1000 proteins were generated per dataset with 10 unmodified features per protein. One PTM was created for each protein in the dataset, resulting in 1000 PTMs, each with 10 modified features. 500 of the PTMs were generated with a differential fold change between conditions, while the other half were generated with no difference between conditions. Of the 500 differential PTMs, the fold change of 250 were fully masked by changes in the unmodified protein. Of the 500 non-differential PTMs, 250 were generated with a differential fold change that was entirely due to the unmodified protein. All differential PTMs were generated with an expected log fold change of .75 between conditions. Each simulation was generated with random biological variation. Specifically, let $Y_{ibc}$ denote the abundance of peptide $i$ in biological replicate $b$ of condition $c$. The simulated peptide abundance  $Z_{ibc}$ was generated as $Z_{ibc} = Y_{ibc} + \epsilon_{ibc}$ where $\epsilon_{ibc}\sim^{iid}N(0,\sigma^2)$. The simulations were generated with one of two values $\sigma^2 = \{.2, .3\}$ motivated by the biological experiments in this manuscript.
Further details can be found in Supplementary Sec. \ref{supp-sec:dataset1}.

\todo{generated (how many?)  repeated synthetic datasets?}. \todo{I have a hard time following this section. Would it be possible to rewrite the description, by explaining (1) how many repeated datasets, (2) how many proteins per dataset; (3) how many PTMs per protein, (4) how many modified/unmodified features per protein? Then talk about the quantitative aspects: systematic fold change; random variation; where the parameters of the simulation were inspired from. It will help relate this simulation to real data} had 1000 PTMs and 1000 unmodified proteins, each represented by 10 LC-MS features observed across all the mass spectrometry runs.

@@

The data were simulated as above, while including missing values and a low number of modified features to provide a more realistic representation of the experiments. In this simulation, the feature counts and the proportion of missing values were as observed on average over all the the experimental datasets in this manuscript. PTMs were simulated with 2 modified peptide features, while unmodified proteins were simulated with 10 features. Additionally, 20\% of observations for both modified and unmodified peptides were missing completely at random. \todo{There will likely be a comment that missing values are more common among low-abundance features. Did the simulation account for that?} Further details can be found in Supplementary Sec. \ref{supp-sec:dataset2}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The proposed approach corrected for high FDR in computer simulations}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{The proposed approach outperformed existing methods in simulations with missing values and low features}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{In a label-free benchmark experiment the proposed approach corrected for bias and outperformed existing methods}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{MSstatsPTM corrected for bias in group comparison TMT experiment}

@@

\subsubsection*{Proposed approach removed confounding in time series TMT experiment}

@@

\subsubsection*{Noisy PTM measurements benefited from additional biological replicates}
